---
title: "Some words about reproducible research"
author: "Vittorio Perduca (Univ. Paris Descartes)"
date: "UiT Tromsø, 21 June 2016"
output: ioslides_presentation
---

## Definition

* _Reproducibility is the ability of an entire experiment or study to be duplicated, either by the same researcher or by someone else working independently._ (Wikipedia)
* Reproducibility is enhanced if analysis (including source code) and results are saved in well documented, transparent and portable supports so that:
    - anyone can rerun the original analysis
    - authors do not get drown in a sea of different versions, trying to match the latest script file with the latest text version.

## Tools for research with statistical contents

* RStudio: authoring contents with text and R output  
    - R Markdown: reports in word, html, pdf (latex-style) and presentations in html
    - R presentations
    - Sweave: reports and presentations on pdf (latex)
    - Shiny: interactive widgets 
* Jupyter and Python notebook: interactive notebooks
* Git and Github: control version, allows to work on same files with collaborators
* Overleaf: collaborative writing



## R Markdown

* Not a WYSIWYG (_what you see is what you get_) authoring system such as MS word and Powerpoint: what you type is _almost_ what you get 
* Easy syntax for writing text paragraphs and tables (though a bit simplistic)
* Allows LaTeX syntax for mathematical formulas
* Easy to embed R code 'chunks' to render plots and output data summary.
* The output file is automatically updated when code and/or data change. 


## Example: slide with R output
```{r}
summary(pressure)
```

## Example: slide with R output 

```{r}
plot(pressure)
```

## Learn R Markdown!

* Many tutorials can be found online. Search a combination of 'R Markdown', 'reproducible research', 'github'...
* For instance: 
    - https://rpubs.com/marschmi/105639
      - R.D. Peng's Mooc: https://www.coursera.org/learn/reproducible-research

